Evaluating infectious disease forecasts with allocation scoring rules

### 2023-11-29 alloscore_manuscript/alloscore-application.Rnw
pass at updating application.

alloscore_models <- unique(alloscore_data_summaries$model)
score_data_summaries <- score_data |>
  mutate(reference_date = target_end_date - 14) |>
  filter(model %in% alloscore_models) |>
  group_by(model, reference_date) |>
  summarize(mwis = mean(wis)) |>
  ungroup() |>
  group_by(reference_date) |>
  mutate(nmodels = n(),
         mwis_rank = rank(mwis, ties.method = "min"),
         mwis_rank_std = 1- (mwis_rank-1)/(nmodels-1)) |>
  ungroup() |>
  complete(model, reference_date) |>
  mutate(target_end_date = as.Date(reference_date) + 14)
@

@@

<<metrics-correlation, fig.width=10, fig.height=7.5, fig.cap="Association of standardized ranks for MWIS and allocation score by model and week. Each facet of the plot corresponds to one model. Within each facet, each point corresponds to a week. The x- and y-values correspond to the MWIS standardized rank and the allocation score standardized rank for that week. Points corresponding to earlier dates have darker shading. The size of the point corresponds to the observed value on the date for which the prediction was made. Models show different degrees of association between the two metrics.">>=
all_data <- left_join(alloscore_data_summaries, score_data_summaries,
                      by=c("model", "target_end_date"))

all_data |>
  filter(as.Date(reference_date.x) < as.Date("2022-02-01") ) |>
  ggplot(aes(x=mwis_rank_std, y=allo_rank_std, color=target_end_date, size=yobs)) +
  geom_point(alpha=.7) +
  scale_x_continuous("MWIS standardized rank", limits=c(0,1)) +
  scale_y_continuous("Allocation score standardized rank", limits=c(0,1)) +
  facet_wrap(.~model) +
  geom_abline(slope=1, intercept=0, linetype=2, color="gray")

@@

@@

\ngr{include figure here that shows relative rankings of locations?}

>>>>

alloscore_models <- unique(alloscore_data_summaries$model)
score_data_summaries <- score_data |>
  mutate(reference_date = target_end_date - 14) |>
  filter(model %in% alloscore_models) |>
  group_by(model, reference_date) |>
  summarize(mwis = mean(wis)) |>
  ungroup() |>
  group_by(reference_date) |>
  mutate(nmodels = n(),
         mwis_rank = rank(mwis, ties.method = "min"),
         mwis_rank_std = 1- (mwis_rank-1)/(nmodels-1)) |>
  ungroup() |>
  complete(model, reference_date) |>
  mutate(target_end_date = as.Date(reference_date) + 14)

all_scores <- left_join(alloscore_data_summaries, score_data_summaries,
                      by=c("model", "target_end_date"))
@

\subsubsection{Anatomy of forecast scores for one week}

To illustrate the mechanics of allocation scoring, we start by focusing on how forecasts generated on or before December 20, 2021, with predictions for January 3, 2022, were scored by different metrics.
This week was around the Omicon wave peak nationally, with individual states typically observing a peak at or after January 3, 2022.

For many locations, forecasts were \ngr{fill in details here}.
On this date, of the 10 models evaluated, the CU-select model had the most accurate forecasts according to the allocation score while the USC-SI\_kJalpha model had the most accurate forecasts based on MWIS (Table \ref{tab:one-week-results}).
The JHUAPL-Bucky model had the second best MWIS but the third worst allocation score.

<<metrics-rankings, results = "asis">>=
## for highlighting cells, if we want that https://stackoverflow.com/questions/37448348/color-code-cells-of-xtable
all_scores |>
  filter(target_end_date == "2022-01-03", !is.na(mwis)) |>
  select(model, alloscore, allo_rank, mwis, mwis_rank) |>
  arrange(allo_rank) |>
  xtable::xtable(label = "tab:one-week-results",
                 caption = "caption TBD") |>
  print(include.rownames = FALSE)
@

Forecasts and allocations for the nine states with the highest numbers of hospitalizations on January 3, 2022 reveal mechanics about how allocations are made (Figure \ref{fig:thermometer-plot}).
\ngr{Aaron - can you add this file to the Sweave doc? Maybe just make a target for it so the figure is generated outside the latex?}
As described in the methods above, allocations for a given location from a specific model are generated by finding the single quantile of the probabilistic forecast across all locations that reaches the resource allocation limit (in this case, assumed to be 15,000).
A direct comparison between the forecasts from the JHUAPL-Bucky and USC-SI\_kJalpha models show that while the JHUAPL-Bucky forecast distributions were closer to the truth across all states, the allocations suggested by those forecasts often were more inefficient than those from the USC-SI\_kJalpha model (Figure \ref{fig:thermometer-plot}).

<<multi-loc-ranks>>=
## date points to target_end_date of 2022-01-10 to highlight
ref_date_for_plot <- "2021-12-20"
models_to_plot <- c("JHUAPL-Bucky", "CU-select")

# tar_load(forecast_data)
# forecast_dat_to_merge <- forecast_data |>
#   dplyr::filter(reference_date == ref_date_for_plot,
#          model %in% c("JHUAPL-Bucky", "MUNI-ARIMA"),
#          quantile == "0.5") |>
#   select(model, reference_date, state = abbreviation, quantile, median = value) |>
#   mutate(reference_date = as.character(reference_date))

multiloc_dat <- Kat15k_alloscores |>
  filter(reference_date == ref_date_for_plot,
         ## the above date points to target_end_dateof 2022-01-10 where things were really wrong
         model %in% models_to_plot) |>
  rename(allocation = x, observation = y, state = abbreviation) |>
  select(model, reference_date, state, allocation, observation) |>
#  left_join(forecast_dat_to_merge) |>
#  pivot_longer(cols = c("allocation", "observation", "median"))
  pivot_longer(cols = c("allocation", "observation"))

## plot all locations observations and allocations
multiloc_plot <- ggplot(multiloc_dat, aes(x=name, y=value, group=state)) +
  geom_line() +
  geom_line(data=dplyr::filter(multiloc_dat, state=="FL"), color="orange") +
  ylab("Daily hospital admissions") +
  scale_x_discrete(name=NULL, limits=rev, expand = c(.1,.1)) +
  facet_grid(.~model)

oracle_rows <- Kat15k_alloscores |>
  filter(reference_date == ref_date_for_plot,
         model %in% c("JHUAPL-Bucky")) |>
  rename(observation = y, state = abbreviation, allocation = oracle) |>
  select(reference_date, state, allocation, observation) |>
  pivot_longer(cols = c("allocation", "observation")) |>
  mutate(model = "Oracle")

multiloc_dat |>
  bind_rows(oracle_rows) |>
  mutate(state = reorder(state, -ifelse(name =="observation", value, 0)),
         name = factor(name, levels = c("observation", "allocation"))) |>
  arrange(name) |>
  ggplot(aes(x=state, y=value, fill=name)) +
  scale_fill_manual(values = c("#CD2626", "goldenrod"))+
  geom_col(position="identity", alpha=.7) +
  facet_grid(model~.)
@

@@

<<metrics-correlation, fig.width=10, fig.height=7.5, fig.cap="Association of standardized ranks for MWIS and allocation score by model and week. Each facet of the plot corresponds to one model. Within each facet, each point corresponds to a week. The x- and y-values correspond to the MWIS standardized rank and the allocation score standardized rank for that week. Points corresponding to earlier dates have darker shading. The size of the point corresponds to the observed value on the date for which the prediction was made. Models show different degrees of association between the two metrics.">>=
all_scores |>
  filter(as.Date(reference_date.x) < as.Date("2022-02-01") ) |>
  ggplot(aes(x=mwis_rank_std, y=allo_rank_std, color=target_end_date, size=yobs)) +
  geom_point(alpha=.7) +
  scale_x_continuous("MWIS standardized rank", limits=c(0,1)) +
  scale_y_continuous("Allocation score standardized rank", limits=c(0,1)) +
  facet_wrap(.~model) +
  geom_abline(slope=1, intercept=0, linetype=2, color="gray")

@@

Models that were able to provide more accurate relative rankings of locations according to the hospitalization counts at the future horizon in general showed better performance according to the allocation score (Figure \ref{fig:multi-loc-ranks}.

@@
